\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
last review of paper

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician is their patient's risk of an event. However, methods
relying on time matching or risk-set sampling (including Cox regression)
eliminate the baseline hazard from the estimating function. The baseline
hazard then needs to be estimated separately using a non-parametric
approach. This leads to step-wise estimates of the survival function
that are difficult to interpret. Using case-base sampling, Hanley \&
Miettinen (2009) explained how parametric hazard functions can be
estimated using logistic regression. Their approach naturally leads to
estimates of the survival function that are smooth-in-time.\\
In this paper, we present the \CRANpkg{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis. We
illustrate all these features through three different case studies.\\
* SRB and MT contributed equally to this work.
}

@@

The semiparametric Cox model has become the default approach to survival
analysis even though Cox himself later suggested he would prefer to
model the hazard function directly. In a 1994 interview with Professor
Nancy Reid, Sir David Cox was asked how he would model a set of censored
survival data, to which he responded: ``I think I would normally want to
tackle problems parametrically \ldots{} and if you want to do things
like predict the outcome for a particular patient, it's much more
convenient to do that parametrically'' \citep{reid1994conversation}.
Indeed, the most relevant quantity in a clinical setting is often the 5-
or 10-year risk of experiencing a certain event given the patient's
particular circumstances. Unfortunately, the most reported metric from a
Cox model is the (potentially time-dependent) hazard ratio (HR), which
ignores the duration of follow-up or is subject to selection bias
\citep{hernan2010hazards}. The covariate adjusted survival curve
overcomes these limitations, and is arguably a more important summary
measure to report than the HR. While stepwise survival curves can be
computed with the Cox model, they require a second step to separately
estimate the baseline hazard \citep{breslow1972discussion}.

Several authors have since pursued fully parametric approaches that made
the fitting of smooth hazard functions much more transparent and
intuitive through generalized linear models. The key feature of these
procedures is splitting the time axis into discrete intervals.
\citep{whitehead1980fitting} showed the equivalence between a Cox model
and a Poisson regression with a parameter for each event time.
\citep{copenhagen2012needs} provides a nice exposition of this
equivalence with a real data example and supporting R code for computing
standard errors. \citep{arjas1987logistic} and \citep{efron1988logistic}
treated each patient-day as a Bernoulli random variable with probability
equal to the discrete hazard rate. A potential issue with these
approaches is that the number of time bins need to be chosen by the data
analyst. A fine grouping of times may result in few (or none) events per
interval, which then leads to instability in the Newton-Raphson
procedure for estimation \citep{kalbfleisch2011statistical}. A coarse
grouping could potentially mask nonlinear trends in the hazard function.

Rather than discretizing time, Hanley \& Miettinen
\citeyearpar{hanley2009fitting} selected a discrete set of person-time
coordinates (`person-moments') in continuous time from all observed
follow-up experience constituting the study base to obtain the logistic
form likelihood expression for the hazard function. More specifically,
all person-moments when the event of interest occurred are selected as
the case series, complemented by a randomly sampled base series of
person-moments serving as controls. This approach allows flexible
modeling of the hazard function by including time as a covariate
(e.g.~using splines or general additive models). Furthermore,
time-dependent covariates can be modeled through interactions with time.
In short, Hanley \& Miettinen \citeyearpar{hanley2009fitting} use the
well-understood logistic regression for directly modeling the hazard
function, but do not require an arbitrary choice of splitting time.

In this article, we present the \CRANpkg{casebase} R package
\citep{casebase-package} which implements the Hanley \& Miettinen
\citeyearpar{hanley2009fitting} approach for fitting fully parametric
hazard models and covariate adjusted survival curves using the familiar
interface of the \code{glm} function. Our implementation allows for
straightforward extensions to other models such as penalized regression
for variable selection and competing-risk analysis. In addition, we
provide functions for exploratory data analysis and visualizing the
estimated quantities such as the hazard function, survival curve and its
standard errors. The ultimate goal of our package is to make fitting
flexible hazards accessible to more endusers with the hopes that they
will favor reporting absolute risks over hazard ratios.

In what follows, we first recall some theoretical details on case-base
sampling and its use for estimating parametric hazard functions. We then
give a short review of existing R packages that implement comparable
features as \pkg{casebase}. Next, we provide some details about the
implementation of case-base sampling in our package, and we give a brief
survey of its main functions. This is followed by three case studies
that illustrate the flexibility and capabilities of \pkg{casebase}. We
show how the same framework can be used for non-linear functions of time
and non-proportional hazards, competing risks and variable selection via
penalized regression. Finally, we end the article with a discussion of
the results and of future directions.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to consider the entire study base
as an infinite collection of \emph{person moments}. These person moments
are indexed by both an individual in the study and a time point, and
therefore each person moment has a covariate profile, an exposure
status, and an outcome status attached to it. By comparing person
moments at which an event occurred with person moments at which no event
occurred, we can extract information about hazards and survival.

@@

To complement the case series, we sample person moments for the base
series. To do so, we model the base series sampling mechanism using
non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
the person-moments where \(dR_i(t) = 1\) constitute the base series. The
process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and
base series person-moments contributed by individual \(i\). The
processes \(R_i(t)\) are typically defined by the user via its intensity
function \(\rho_i(t)\). The process \(Q_{i}(t)\) is characterized by
\(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).

If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
terms of \(\theta\), we could define an estimator \(\hat{\theta}\) by
maximization of the likelihood expression
\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},\]
where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
\(u\), and where \(t_i\) is the event time for individual \(i\).
However, the integral over time makes the computation and maximization
of \(L_0(\theta)\) challenging.

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
\end{equation}

In this model, the instantaneous failure rate is independent of
\(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

Case-base sampling can also be used in the context of competing-risk
analyses. Assuming there are \(J\) competing events, we can show that
each person-moment's contribution to the likelihood is of the form

\[\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]

where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(\lambda_j(t)\) is the corresponding cause-specific
hazard function. As may be expected, this functional form is similar to
the terms appearing in the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

At the time of writing, a cursory examination of CRAN's \ctv{Survival}
Task View reveals that there are over 250 packages related to survival
analysis \citep{survTaskView}. For the purposes of this article, we
restricted our review to packages that implement at least one of the
following features: parametric modeling, non-proportional hazard models,
competing risk analysis, penalized estimation, and CIF estimation. By
searching for appropriate keywords in the \code{DESCRIPTION} file of
these packages, we found 60 relevant packages. These 60 packages were
then manually examined to determine which ones are comparable to
\pkg{casebase}. In particular, we excluded packages that were focused on
a different set of problems, such as frailty and multi-state models. The
remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with
some of the functionalities they offer.

Parametric survival models are implemented in a handful of packages,
each differing in the underlying data distribution: \CRANpkg{CFC}
\citeyearpar{mahani2015bayesian}, \CRANpkg{flexsurv}
\citeyearpar{flexsurv}, \CRANpkg{SmoothHazard}
\citeyearpar{smoothHazard}, \CRANpkg{rstpm2} \citeyearpar{clements_liu},
\CRANpkg{mets} \citeyearpar{scheike2014estimating}, and
\CRANpkg{survival} \citeyearpar{survival-package}. For example,
\pkg{SmoothHazard} is limited to Weibull distributions
\citeyearpar{smoothHazard}, whereas both \pkg{flexsurv} and
\pkg{survival} allow users to supply any distribution of their choice.
\pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can
model the effect of time using splines, which allows flexible modeling
of the hazard function. As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear
combination of covariates (including time). Therefore, our package is
more general in that it allows the user to model any linear or
non-linear transformation of time including splines and higher order
polynomials. Also, by including interaction terms between covariates and
time, it also allows users to fit (non-proportional) time-varying
coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly
model any shape parameter.

@@

Next, several R packages implement methodologies for competing risk
analysis; for a different perspective on this topic, see Mahani \&
Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{survival}
provides functionality for competing-risk analysis and multistate
modelling. The package \CRANpkg{cmprsk} provides methods for
cause-specific subdistribution hazards, such as in the Fine-Gray model
\citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric
survival functions. Our package \pkg{casebase} also provides
functionalities for competing risk analysis by estimating parametrically
the cause-specific hazards. From these quantities, we can then estimate
the cause-specific CIFs.

Finally, several packages include functions to estimate the survival
function and the CIF. The corresponding methods generally fall into two
categories: transformation of the estimated hazard function, and
semi-parametric estimation of the baseline hazard. The first category
broadly corresponds to parametric survival models, where the full hazard
is explicitly modeled. Using this estimate, the survival function and
the CIF can be obtained using their functional relationships (see
Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing
this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and
\pkg{survival}. Our package \pkg{casebase} also follows this approach
for both single-event and competing-risk analyses. The second category
outlined above broadly corresponds to semi-parametric models. These
models do not model the full hazard function, and therefore the baseline
hazard needs to be estimated separately in order to estimate the
survival function. This is achieved using semi-parametric estimators
(e.g.~Breslow's estimator) or parametric estimators (e.g.~spline
functions). Packages that implement this approach include
\pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As
mentioned in the introduction, a key distinguishing factor between these
two approaches is that the first category leads to smooth estimates of
the survival function, whereas the second category often produces
estimates in the form of stepwise functions.

@@

\begin{Schunk}
\begin{Sinput}
pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                               event = "DeadOfPrCa", exposure = "ScrArm")
inherits(pt_object, "popTime")
\end{Sinput}
\begin{Soutput}
#> [1] TRUE
\end{Soutput}
\begin{Sinput}
attr(pt_object, "exposure")
\end{Sinput}
\begin{Soutput}
#> [1] "ScrArm"
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./plot-stratified-erspc-data-1} \caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments). More events are observed at later follow-up times, motivating the use of non-constant hazard models.}\label{fig:plot-stratified-erspc-data}
\end{figure}
\end{Schunk}

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by
first adding a layer for the area representing the population time in
gray, with subjects having the least amount of observation time plotted
at the top of the y-axis. We immediately notice a distinctive
\emph{stepwise shape} in the population time area. This is due to the
randomization of the Finnish cohorts which were carried out on January 1
of each of year from 1996 to 1999. Coupled with the uniform December 31
2006 censoring date, this led to large numbers of men with exactly 11,
10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from
right to left), the population-time plot shows the recruitment pattern
from its beginning in 1991, and the January 1 entries in successive
years. Tracked forwards in time (i.e.~from left to right), the plot for
the first three years shows attrition due entirely to death (mainly from
other causes). Since the Swedish and Belgian centres were the last to
complete recruitment in December 2003, the minimum potential follow-up
is three years. Tracked further forwards in time (i.e.~after year 3) the
attrition is a combination of deaths and staggered entries. As we can
see, population-time plots summarise a wealth of information about the
study into a simple graph.

Next, layers for the case series and base series are added. The y-axis
location of each case moment is sampled at random vertically on the plot
to avoid having all points along the upper edge of the gray area. By
randomly distributing the cases, we can get a sense of the incidence
density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that
more events are observed at later follow-up times. Therefore, a constant
hazard model would not be appropriate in this instance as it would
overestimate the incidence earlier on in time, and underestimate it
later on. Finally, the base series is sampled horizontally with sampling
weight proportional to their follow-up time. The reader should refer to
the package vignettes for more examples and a detailed description of
how to modify the aesthetics of a population-time plot.

@@

Finally, by integrating these subdensities, we obtain the event-specific
cumulative incidence functions:

@@

@@

In this section, we illustrate the main functions of the \pkg{casebase}
package through three case studies. Each one showcases a different type
of analysis. First, we show how to model different functional forms of
time as well as non-proportional hazards. Then we perform a
competing-risk analysis and compare our results with the Fine-Gray
model. The third case study illustrates how to perform variable
selection in high-dimensional datasets.

@@

Moreover, we can also compute confidence intervals for the risk function
using the method \code{confint.absRiskCB}.

@@

Previous re-analyses of these data suggested that the overall screening
attributed reduction in death due to prostate cancer of 20\% was an
underestimate \citep{hanley2010mortality}. The estimated 20\% (from a
proportional hazards model) did not account for the delay between
screening and the time the effect is expected to be observed. As a
result, the null effects in years 1--7 masked the substantial reductions
that began to appear from year 8 onward. This motivates the use of a
time-dependent hazard ratio which can easily be fit with the
\pkg{casebase} package by including an interaction term with time in the
model:

@@

In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard
ratio and 95\% confidence interval for screening vs.~control group as a
function of time using the \code{plot} method for objects of class
\code{singleEventCB}:

@@

The plot shows that the cures attributable to the screening only begin
to become statistically apparent by year 7 and later. The 25-60\%
reductions seen in years 8-12 of the study suggests a much higher
reduction in prostate cancer due to screening than the single overall
20\% reported in the original article.

@@

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.21)\\
Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.28, 0.94)\\
Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.47, 3.09)\\
Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.46, 6.08)\\
Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (1.98, 10.46)\\
\addlinespace
Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.53)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

From the fit object, we can extract both the hazard ratios and their
corresponding confidence intervals. These quantities appear in Table
\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
identified by case-base sampling is the one associated with the phase of
the disease at transplant. More precisely, being in relapse at
transplant is associated with a hazard ratio of 3.89 when compared to
CR1.

>>>>

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician is their patient's risk of an event. However, methods
relying on time matching or risk-set sampling (including Cox regression)
eliminate the baseline hazard from the estimating function. As a
consequence, the focus has been on reporting hazard ratios instead of
survival curves. Indeed, reporting patient risk requires a separate
estimation of the baseline hazard. Using case-base sampling, Hanley \&
Miettinen (2009) explained how parametric hazard functions can be
estimated in continuous-time using logistic regression. Their approach
naturally leads to estimates of the survival function that are
smooth-in-time.\\
In this paper, we present the \CRANpkg{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis. We
illustrate all these features through three different case studies.\\
* SRB and MT contributed equally to this work.
}

@@

The semiparametric Cox model has become the default approach to survival
analysis even though Cox himself later suggested he would prefer to
model the hazard function directly. In a 1994 interview with Professor
Nancy Reid, Sir David Cox was asked how he would model a set of censored
survival data, to which he responded: ``I think I would normally want to
tackle problems parametrically \ldots{} and if you want to do things
like predict the outcome for a particular patient, it's much more
convenient to do that parametrically'' \citep{reid1994conversation}.
Indeed, the most relevant quantity in a clinical setting is often the 5-
or 10-year risk of experiencing a certain event given the patient's
particular circumstances. Unfortunately, the most reported metric from a
Cox model is the (potentially time-dependent) hazard ratio (HR), which
ignores the duration of follow-up and is subject to selection bias
\citep{hernan2010hazards}. The covariate-adjusted survival curve
overcomes these limitations, and it is arguably a more important summary
measure to report than the HR. While stepwise survival curves can be
computed with the Cox model, they require a second step to separately
estimate the baseline hazard \citep{breslow1972discussion}.

Several authors have since pursued fully parametric approaches that made
the fitting of smooth survival curves much more transparent and
intuitive through generalized linear models. The key feature of these
procedures is splitting the time axis into discrete intervals. Whitehead
\citeyearpar{whitehead1980fitting} showed the equivalence between a Cox
model and a Poisson regression with a parameter for each event time;
Carstensen \citeyearpar{copenhagen2012needs} provides a nice exposition
of this equivalence with a real data example and supporting R code for
computing standard errors. Arjas \& Haara
\citeyearpar{arjas1987logistic} and Efron
\citeyearpar{efron1988logistic} treated each patient-day as a Bernoulli
random variable with probability equal to the discrete hazard rate. A
potential issue with these approaches is that the number of time bins
need to be chosen by the data analyst. On the one hand, a fine grouping
of times may result in few (or none) events per interval, which then
leads to instability in the Newton-Raphson procedure for estimation
\citep{kalbfleisch2011statistical}. On the other hand, a coarse grouping
could potentially mask nonlinear trends in the hazard function.

Rather than discretizing time, Hanley \& Miettinen
\citeyearpar{hanley2009fitting} selected a discrete set of person-time
coordinates (`person-moments') in continuous time from all observed
follow-up experience constituting the study base. By doing so, they
obtained a likelihood expression for the hazard function that is
equivalent to that of logistic regression. More specifically, all
person-moments when the event of interest occurred are selected as the
case series, complemented by a randomly sampled base series of
person-moments serving as controls. This approach allows flexible
modeling of the hazard function by including time as a covariate
(e.g.~using splines or general additive models). Furthermore,
time-dependent covariates can be modeled through interactions with time.
In short, Hanley \& Miettinen \citeyearpar{hanley2009fitting} use the
well-understood logistic regression for directly modeling the hazard
function, without requiring a discrete-time model.

In this article, we present the \CRANpkg{casebase} R package
\citep{casebase-package} which implements the Hanley \& Miettinen
\citeyearpar{hanley2009fitting} approach for fitting fully parametric
hazard models and covariate-adjusted survival curves using the familiar
interface of the \code{glm} function. Our implementation allows for
straightforward extensions to other models such as penalized regression
for variable selection and competing-risk analysis. In addition, we
provide functions for exploratory data analysis and visualizing the
estimated quantities such as the hazard function, survival curve, and
their standard errors. The ultimate goal of our package is to make
fitting flexible hazards accessible to more end users with the hope that
they will favor reporting absolute risks over hazard ratios.

In what follows, we first recall some theoretical details on case-base
sampling and its use for estimating parametric hazard functions. We then
give a short review of existing R packages that implement comparable
features as \pkg{casebase}. Next, we provide some details about the
implementation of case-base sampling in our package, and we give a brief
survey of its main functions. This is followed by three case studies
that illustrate the flexibility and capabilities of \pkg{casebase}. We
show how the same framework can be used for non-linear functions of time
and non-proportional hazards, competing risks, and variable selection
via penalized regression. Finally, we end the article with a discussion
of the results and of future directions.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to consider the entire study base
as an infinite collection of \emph{person moments}. These person moments
are indexed by both an individual in the study and a time point, and
therefore each person moment has a covariate profile, an exposure
status, and an outcome status (i.e.~whether the event happened) attached
to it. By comparing person moments at which an event occurred with
person moments at which no event occurred, we can extract information
about hazards and survival.

@@

To complement the case series, we sample person moments for the base
series. To do so, we model the base series sampling mechanism using
non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
the person-moments where \(dR_i(t) = 1\) constitute the base series. We
note that the same individual can contribute multiple person-moments to
the base series. The process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then
counts both the case and base series person-moments contributed by
individual \(i\). As mentioned above, the processes \(R_i(t)\) are
typically defined by the user via its intensity function \(\rho_i(t)\).
The process \(Q_{i}(t)\) is characterized by
\(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).

If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
terms of \(\theta\), we can define an estimator \(\hat{\theta}\) by
maximization of the likelihood expression
\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},\]
where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
\(u\), and where \(t_i\) is the event time for individual \(i\).
However, the integral over time makes the computation and maximization
of \(L_0(\theta)\) challenging.

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
\end{equation} In this model, the instantaneous failure rate is
independent of
\(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

Case-base sampling can also be used in the context of competing-risk
analysis. Assuming there are \(J\) competing events, we can show that
each person-moment's contribution to the likelihood is of the form

\[\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(\lambda_j(t)\) is the corresponding cause-specific
hazard function. As may be expected, this functional form is similar to
the terms appearing in the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

At the time of writing, a cursory examination of CRAN's \ctv{Survival}
Task View reveals that there are over 250 packages related to survival
analysis \citep{survTaskView}. For the purposes of this article, we
restricted our review to packages that implement at least one of the
following features: parametric modeling, non-proportional hazard models,
competing risk analysis, penalized estimation, and Cumulative Incidence
(CI) estimation. By searching for appropriate keywords in the
\code{DESCRIPTION} file of these packages, we found 60 relevant
packages. These 60 packages were then manually examined to determine
which ones are comparable to \pkg{casebase}. In particular, we excluded
packages that were focused on a different set of problems, such as
frailty and multistate models. The remaining 14 packages appear in Table
\ref{tab:surv-pkgs}, along with some of the functionalities they offer.

Parametric survival models are implemented in several packages, each
differing in the parametric distributions available: \CRANpkg{CFC}
\citeyearpar{mahani2015bayesian}, \CRANpkg{flexsurv}
\citeyearpar{flexsurv}, \CRANpkg{SmoothHazard}
\citeyearpar{smoothHazard}, \CRANpkg{rstpm2} \citeyearpar{clements_liu},
\CRANpkg{mets} \citeyearpar{scheike2014estimating}, and
\CRANpkg{survival} \citeyearpar{survival-package}. For example,
\pkg{SmoothHazard} is limited to Weibull distributions
\citeyearpar{smoothHazard}, whereas both \pkg{flexsurv} and
\pkg{survival} allow users to supply any distribution of their choice.
\pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can
model the effect of time using splines, which allows flexible modeling
of the hazard function. As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear
combination of covariates (including time). Therefore, our package is
more general in that it allows the user to model any linear or
non-linear transformation of time including splines and higher order
polynomials. Also, by including interaction terms between covariates and
time, it also allows users to fit (non-proportional) time-varying
coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly
model any shape parameter.

@@

Next, several R packages implement methodologies for competing risk
analysis; for a different perspective on this topic, see Mahani \&
Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{survival}
provides functionality for competing-risk analysis and multistate
modelling. The package \CRANpkg{cmprsk} provides methods for
cause-specific subdistribution hazards, such as in the Fine-Gray model
\citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric
survival functions. Our package \pkg{casebase} also provides
functionalities for competing risk analysis by estimating parametrically
the cause-specific hazards. From these quantities, we can then estimate
the cause-specific CIs.

Finally, several packages include functions to estimate the survival
function and the CI. The corresponding methods generally fall into two
categories: transformation of the estimated hazard function, and
semi-parametric estimation of the baseline hazard. The first category
broadly corresponds to parametric survival models, where the full hazard
is explicitly modeled. Using this estimate, the survival function and
the CI can be obtained using their functional relationships (see
Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing
this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and
\pkg{survival}. Our package \pkg{casebase} also follows this approach
for both single-event and competing-risk analyses. The second category
outlined above broadly corresponds to semi-parametric models. These
models do not model the full hazard function, and therefore the baseline
hazard needs to be estimated separately in order to estimate the
survival function. This is achieved using semi-parametric estimators
(e.g.~Breslow's estimator) or parametric estimators (e.g.~spline
functions). Packages that implement this approach include
\pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As
mentioned in the introduction, a key distinguishing factor between these
two approaches is that the first category leads to smooth estimates of
the survival function, whereas the second category often produces
estimates in the form of stepwise functions.

@@

\begin{Schunk}
\begin{Soutput}
#> [1] TRUE
\end{Soutput}
\begin{Soutput}
#> [1] "ScrArm"
\end{Soutput}
\end{Schunk}

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./plot-erspc-data-1} \caption{Population time plot for the ERSPC dataset. \textbf{A}: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up. \textbf{B}: The red points correspond to when death has occurred for any one of those infinitely thin rectangles. \textbf{C}: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density. More events are observed at later follow-up times, motivating the use of non-constant hazard models. \textbf{D}: The base series, a representative sample of the entire grey area, is represented by the green points.}\label{fig:plot-erspc-data}
\end{figure}
\end{Schunk}

@@

Figure \ref{fig:plot-erspc-data} depicts the process of creating a
population-time plot. It is built sequentially by first adding a layer
for the area representing the population time in gray (Figure
\ref{fig:plot-erspc-data} A), with subjects having the least amount of
observation time plotted at the top of the y-axis. We immediately notice
a distinctive \emph{stepwise shape} in the population time area. This is
due to the randomization of the Finnish cohorts which were carried out
on January 1 of each of year from 1996 to 1999. Coupled with the uniform
December 31 2006 censoring date, this led to large numbers of men with
exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time
(i.e.~from right to left), the population-time plot shows the
recruitment pattern from its beginning in 1991, and the January 1
entries in successive years. Tracked forwards in time (i.e.~from left to
right), the plot for the first three years shows attrition due entirely
to death (mainly from other causes). Since the Swedish and Belgian
centres were the last to complete recruitment in December 2003, the
minimum potential follow-up is three years. Tracked further forwards in
time (i.e.~after year 3) the attrition is a combination of deaths and
staggered entries. As we can see, population-time plots summarise a
wealth of information about the study into a simple graph.

Next, layers for the case series and base series are added. The y-axis
location of each case moment is sampled at random vertically on the plot
to avoid having all points along the upper edge of the gray area (Figure
\ref{fig:plot-erspc-data} B). By randomly distributing the cases, we can
get a sense of the incidence density. In Figure
\ref{fig:plot-erspc-data} C, we see that more events are observed at
later follow-up times. Therefore, a constant hazard model would not be
appropriate in this instance as it would overestimate the incidence
earlier on in time, and underestimate it later on. Finally, the base
series is sampled uniformly from the study base (Figure
\ref{fig:plot-erspc-data} D). The reader should refer to the package
vignettes for more examples and a detailed description of how to modify
the aesthetics of a population-time plot.

@@

By integrating these subdensities, we obtain the event-specific CI
functions:

@@

Finally, the output from \code{absoluteRisk} can be passed to a method
\code{confint} to compute confidence bands around the survival function.
These bands are computed using parametric bootstrap, and therefore are
only valid when \code{family = "glm"} as it relies on the asymptotic
normality of the estimator. Currently, this is only available for the
single-event setting.

@@

In this section, we illustrate the main functions of the \pkg{casebase}
package through three case studies. Each one showcases a different type
of analysis. First, we show how to model time flexibly as well as
non-proportional hazards. Then we perform a competing-risk analysis and
compare our results with the Cox model and the Fine-Gray model. The
third case study illustrates how to perform variable selection in
high-dimensional datasets.

@@

Next, we can compute confidence intervals for the risk function using
the method \code{confint.absRiskCB}.

@@

Previous re-analyses of these data suggest that the 20\% reduction in
prostate cancer death due to screening was an underestimate
\citep{hanley2010mortality}. The estimated 20\% (from a proportional
hazards model) did not account for the delay between screening and the
time the effect is expected to be observed. As a result, the null
effects in years 1--7 masked the substantial reductions that began to
appear from year 8 onward. This motivates the use of a time-dependent
hazard ratio which can easily be fit with the \pkg{casebase} package by
including an interaction term with time in the model:

@@

Note that we did not have to specify any cut points, as would be the
case with the \code{survSplit} function in the \pkg{survival} package.
In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard
ratio and 95\% confidence interval for screening vs.~control group as a
function of time using the \code{plot} method for objects of class
\code{singleEventCB}:

@@

The plot shows that the effect of screening only becomes statistically
apparent by year 7 and later. The 25-60\% reductions seen in years 8-12
of the study suggests a much higher reduction in prostate cancer due to
screening than the single overall 20\% reported in the original article.

@@

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% Conf. & HR & 95\% Conf.\\
\midrule
Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.21)\\
Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.28, 0.94)\\
Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.47, 3.09)\\
Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.46, 6.08)\\
Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (1.98, 10.46)\\
\addlinespace
Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.53)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

From the fit object, we can extract both the hazard ratios and their
corresponding confidence intervals. These quantities appear in Table
\ref{tab:bmtcrr-cis}. As we can see, the type of disease corresponds to
a significant hazard ratio: the hazard for AML is 48\% lower than that
for ALL. Moreover, being in relapse at transplant is associated with a
hazard ratio of 4.09 when compared to CR1.
